\documentclass{article}
\begin{document}
Pre Oct 8th, 2010
\begin{itemize}
\item Hom(FreeSet(X),L) =~= Hom(X,Forget(L)) where X in Setoid was extremely helpful. Get equality respecting for free, as well as homomorphism. Dan's DFA stuff (fUnion).
\item Universals (Sys F) can encode inductive datatypes, existentials can encode coinductive. Definition of existential in terms of universal in Sys F is neat.
\item "Physics, Logic, Topology and Computation: A Rosetta Stone" is cool. Nice intro to linear logic + category theory (symmetric complete monoidal categories)
\end{itemize}
Oct 8th, 2010
\begin{itemize}
\item Can we get away with just one binder instead of ``multibinder" that g:ctx is? i..e products on the left.
\item Meeting: Have to keep $\psi : \omega$ in the same context $\delta$ as $u::P[\psi]$ because of dependencies. i.e. I think $\omega$ can depend on $u$?
\item Colloqium: Iteratively solving linear systems (Tomograph) means only need small parts of matrix in memory at one time. Makes me think of lazy evaluation and linear types. Here, matrix is a "simple" function of geometry, which is what makes this work. How can different linear types affect peformance?
\end{itemize}
Oct 9th, 2010
\begin{itemize}
\item The HOAS to de Bruijn conversion looks very much like a primitive recursive fold pattern. What else fits this pattern and how can we represent it? (Asn 5)
\item Termination checking in Beluga: The decreasing argument can be $\psi$!
\item Need to try DSLs in Beluga
\item Generics + levitation => encode the logic you want to work in, and lift the results to the metalanguage? Can we, e.g. add some kind of "lifting" axiom tha lets you prove metaresults (parametricity/theorems for free), and bring them into the metalanguage?
\item Cryptography requires epistemic logic. "Adversary cannot compute any function of plaintexts" -- not true, he can compute the encryption function applied to plaintexts, but he doesn't know which function he is computing.

\end{itemize}
\end{document}
